Serum Biomarkers of Nutrition Pre and Post CFTR Modulator Use in Children with Cystic Fibrosis

### 2022-04-15 Edith Zemanick/CF Nutrition/cf_nutrition.Rmd
better CFTR nutrition data cleaning

```{r}
# Set up data again
df = read.csv("./Edith Zemanick/CF Nutrition/Data_Cleaned/analysis_dataset.csv")
df$Age.at.Start..years. = NULL
# New age at start
mods = read_excel("./Edith Zemanick/CF Nutrition/Data_Raw/Report CFTR Modulators_throughDecember2021.xlsx")
mods = mods[mods$Modulator %in% c('Kalydeco','Trikafta'),]
colnames(mods) = make.names(colnames(mods))
mods = mods %>% select(Patient.ID,Modulator,Age.at.Start..years.) %>% 
  pivot_wider(id_cols = Patient.ID,names_from = Modulator,
              values_from = Age.at.Start..years.)
colnames(mods)[2:ncol(mods)] = paste0("age_at_start_",colnames(mods)[2:ncol(mods)])
# Age at first and second modulator
mods$age_first = pmin(mods$age_at_start_Trikafta,mods$age_at_start_Kalydeco,na.rm = T)
mods$age_second = pmax(mods$age_at_start_Trikafta,mods$age_at_start_Kalydeco)
# Merge back in
df = left_join(df,mods,by = "Patient.ID")
df = df %>% group_by(Patient.ID) %>% 
  fill(age_at_start_Trikafta,age_at_start_Kalydeco,.direction = "downup")
# Time relative to first mod
df$time = df$Age.at.Test..Years. - df$age_first
# Calculate spline things
df$c1 = 
  (df$time-0)*(df$time>=0)
df$c2 = 
  (df$Age.at.Test..Years.-df$age_second)*(df$Age.at.Test..Years.>=df$age_second)
# Limit to 3 years prior, select columns
df = df %>% filter(!is.na(time) & time >= -3) %>% 
  select(Patient.ID,Age.at.Test..Years.,age_first,age_second,time,c1,c2,
         all_of(nutrition_vars))
df$c2[is.na(df$c2)] = 0
```

```{r}
piecewise_mod = function(outcome,data,diagnostics = F){
  data$Patient.ID = as.factor(data$Patient.ID)
  data = data %>% select(Patient.ID,time,c1,c2,age_first,age_second,
                         all_of(outcome)) %>% drop_na(all_of(outcome))
  # Separate data for single and double mods
  single = data %>% filter(is.na(age_second))
  double = data %>% filter(!is.na(age_second))
  # Models
  f_single = as.formula(paste0(outcome,"~ time + c1 + (1|Patient.ID)"))
  m_single = lmer(f_single,data = single)
  f_double = as.formula(paste0(outcome,"~ time + c1 +c2 + (1|Patient.ID)"))
  m_double = lmer(f_double,data = double)

  # Diagnostics if required
  if(diagnostics){
    check_single = check_model(m_single)
    check_double = check_model(m_double)
  } else {
    check_single = NULL
    check_double = NULL
  }
  # Results
  res_single = tidy(m_single,"fixed")
  res_double = tidy(m_double,"fixed")
  # Plots
  single_plot_df = m_single@frame
  single_plot_df$pred = predict(m_single)
  single_plot = ggplot(single_plot_df,aes_string(x = "time",y = outcome,color = "Patient.ID")) + 
    geom_line(aes(y = pred)) + 
    theme_bw() + theme(legend.position = "none") + xlab("Time (years)")

  double_plot_df = m_double@frame
  double_plot_df$pred = predict(m_double)
  double_plot = ggplot(double_plot_df,aes_string(x = "time",y = outcome,color = "Patient.ID")) + 
    geom_line(aes(y = pred)) + 
    theme_bw() + theme(legend.position = "none") + xlab("Time (years)")
  # Return
  return(list("results_single" = res_single,"results_double" = res_double,
              "diagnostics_single" = check_single,"diagnostics_double" = check_double,
              "plot_single" = single_plot,"plot_double" = double_plot))
}
```

@@

```{r cache=TRUE,echo=TRUE}
mod_df = df %>% select(Patient.ID,time,VitaminA_Retinol) %>% drop_na()
model = list(
  VitaminA_Retinol ~ 1, # Plateau 
  1 + (1|Patient.ID) ~ 0 + time, # Joined slope
  1 + (1|Patient.ID) ~ 0 + time, # Joined slope
  1 + (1|Patient.ID) ~ 0 + time # Joined slope
)
fit = mcp(model, data = mod_df)
```

```{r}
plot(fit) + theme_bw()
```

>>>>

```{r}
# Set up data again
vit = read_excel("./Edith Zemanick/CF Nutrition/Data_Raw/Report Vitamins_throughDecember2021.xlsx")
vit = vit %>% select(`Patient ID`,Date,VitaminA_Retinol:`Vitamin E Gamma`)
# New age at start
mods = read_excel("./Edith Zemanick/CF Nutrition/Data_Raw/Report CFTR Modulators_throughDecember2021.xlsx")
mods = mods[mods$Modulator %in% c('Kalydeco','Trikafta'),]
mods = mods %>% select(`Patient ID`,Modulator,`Earliest Date in Registry`) %>% 
  pivot_wider(id_cols = `Patient ID`,names_from = Modulator,
              values_from = `Earliest Date in Registry`)
colnames(mods)[2:ncol(mods)] = paste0(colnames(mods)[2:ncol(mods)]," Date")
mods$`First Mod Date` = pmin(mods$`Trikafta Date`,mods$`Kalydeco Date`,na.rm = T)
mods$`Second Mod Date` = pmax(mods$`Trikafta Date`,mods$`Kalydeco Date`)
# Add pancreatic status
pancreatic = read_excel("./CF Info/Sweat Tests from Internal Database_genotype class_withPancStatus.xlsx")
pancreatic = pancreatic %>% rename(`Patient ID` = PatientID) %>% select(`Patient ID`,PancreaticStatus)
# FEV1 and BMI
fev = read_excel("./Edith Zemanick/CF Nutrition/Data_Raw/Report_PFTs_ThroughDecember2021.xlsx")
fev = fev %>% select(PatientID,Date,BMI,`FEV1 % pred`) %>% rename(`Patient ID` = PatientID)
# Merge
df = full_join(vit,fev) %>% filter(`Patient ID` %in% mods$`Patient ID`,)
df = left_join(df,mods)
df = left_join(df,pancreatic)
# Time relative to first mod
df$time = as.numeric(difftime(df$Date,df$`First Mod Date`,units = "days"))/365.25
# Calculate spline things
df$c1 = 
  (df$time-0)*(df$time>=0)
df$c2 = 
  (as.numeric(difftime(df$Date,df$`Second Mod Date`,units = "days"))/365.25)*(df$Date >= df$`Second Mod Date`)
# Limit to 3 years prior, select columns
df = df %>% filter(!is.na(time) & time >= -3) %>% distinct()
# Vitamin A to numeric
df$VitaminA_Retinol = suppressWarnings(as.numeric(df$VitaminA_Retinol))
```

```{r}
piecewise_mod = function(outcome,data,diagnostics = F){
  data$`Patient ID` = as.factor(data$`Patient ID`)
  data = data %>% select(`Patient ID`,Date,`First Mod Date`,`Second Mod Date`,PancreaticStatus,BMI,
                         all_of(outcome)) %>% drop_na(all_of(outcome))
  # Separate data for single and double mods
  single = data %>% filter(c2 <= 0 | is.na(c2))
  double = data %>% filter(!is.na(c2))
  # Models
  f_single = as.formula(paste0(outcome,"~ BMI + PancreaticStatus + time + c1 + (1|`Patient ID`)"))
  f_double = as.formula(paste0(outcome,"~ BMI + PancreaticStatus + time + c1 +c2 + (1|`Patient ID`)"))
  if(outcome == "BMI"){
    f_single = update(f_single,.~.-BMI)
    f_double = update(f_double,.~.-BMI)
  }
  m_single = lmer(f_single,data = single)
  m_double = lmer(f_double,data = double)
  # Diagnostics if required
  if(diagnostics){
    check_single = check_model(m_single)
    check_double = check_model(m_double)
  } else {
    check_single = NULL
    check_double = NULL
  }
  # Results
  res_single = tidy(m_single,"fixed")
  res_double = tidy(m_double,"fixed")
  # Plots
  single_plot_df = m_single@frame
  single_plot_df$pred = predict(m_single)
  single_plot = ggplot(single_plot_df,aes_string(x = "time",y = outcome,color = "`Patient ID`")) + 
    geom_line(aes(y = pred)) + 
    theme_bw() + theme(legend.position = "none") + xlab("Time (years)")
  double_plot_df = m_double@frame
  double_plot_df$pred = predict(m_double)
  double_plot = ggplot(double_plot_df,aes_string(x = "time",y = outcome,color = "`Patient ID`")) + 
    geom_line(aes(y = pred)) + 
    theme_bw() + theme(legend.position = "none") + xlab("Time (years)")
  # Return
  return(list("results_single" = res_single,"results_double" = res_double,
              "diagnostics_single" = check_single,"diagnostics_double" = check_double,
              "plot_single" = single_plot,"plot_double" = double_plot))
}
```

In the following models, the "time" estimate represents the slope prior to first modulator start, "c1" is the change in slope after first modulator start, and "c2" is the change in slope from first modulator to second. All models were adjusted for BMI and pancreatic status (except the BMI model, which was only adjusted for pancreatic status).

@@

```{r cache=TRUE,echo=TRUE}
mod_df = df %>% filter(!is.na(age_second)) %>% select(Patient.ID,time,BMI) %>% drop_na()
model = list(
  BMI ~ 1, # Plateau 
  1 + (1|Patient.ID) ~ 0 + time, # Joined slope
  1 + (1|Patient.ID) ~ 0 + time, # Joined slope
  1 + (1|Patient.ID) ~ 0 + time # Joined slope
)
fit = mcp(model, data = mod_df,sample = "both")
```

### Posterior means

```{r}
summary(fit)
```

### Posterior fit

```{r}
plot(fit,q_fit = TRUE) + theme_bw()
```

### Parameter plot

```{r}
plot_pars(fit)
```

### Hypothesis tests

```{r}
hypothesis(fit,c("cp_1 = 0","cp_2 = 0","cp_3 = 0"))
```
